\section{Introduction}
\label{sec:intro}
At the start and during the COVID-19 pandemic, rising infection rates, deaths and widespread lockdown measures raised questions on how to limit the pandemic's impact on economic development \cite{velavan2020covid}. This thesis looks into how the pandemic affects stock market values, as these are notoriously volatile in their pricing and influenced by a wide range of factors, demonstrated by the WTI value dipping in the negative at the first wave of infections \cite{CORBET2020104978}. 
The relationship between stock markets and economic factors falls under the study of macroeconomics and finds it origins in the business cycle theory, dating back to the early 1800s \cite{de1827nouveaux}. This theory generally states that all economic development follows a cycle or waveform, going through the phases expansion, crisis, recession and recovery. These business cycles can be measured \cite{baxter1999measuring} and are part of the tools that a government can use to mitigate economic downturn. Predicting the development of a business cycle is generally based on economic indicators and economic theory, such as Keynesian theory  \cite{keynes2018general} and real-business cycle (RBC) theory \cite{kydland1982time}. One of the generally accepted economic indicators is the Standard and Poor's (S\&P) 500 as an indicator for the United States (US) economy, stating a relationship between economic development and the top 500 US companies. % \cite{diblasi_2021}
One of the complexities of these relationships, between a country and a national representation of the local stock market, is that a stock market is not a local phenomenon. Stocks and other financial products are traded internationally via a stock exchange and these stock indexes, such as the S\&P500, are a representation of the stock exchange itself and not (necessarily) of the country it resides in. As the pandemic's impact and government restrictions varies between countries, it is needed to isolate a region to perform a causal analysis of its impact on a stock market.
When the WTI value dropped, all oil-related stocks dropped in value as a consequence, such as Royal Dutch Shell, one of the largest companies in the Netherlands. As previous research noted a relationship between pandemic data and economic development, and a relationship between economic development and stock indexes, the question can be posed whether these relationships aid in predictive modelling of stock values in a region. This research will focus on the Netherlands, with the Amsterdam Exchange index (AEX) as an analog for the S\&P500, the former featuring the top 25 most traded securities on the Amsterdam Exchange. As macroeconomics looks at the large scale development of an economy, the models generally apply to years or decades of data. Daily stock market development is a lot more volatile than other economic indicators, generally displaying a non-linear patterns. 
Machine learning can aid in these predictions, as it can capture non-linear behaviour better than traditional linear models \cite{zhang2009stock}. Still, stock market prediction is non-trivial as the efficient market hypothesis (EMH) \cite{peters1996chaos} states that no investor can gain an advantage over others based on historical and current information. Some investors base their decisions on fundamental analysis \cite{lev1993fundamental} to gain an advantage on others. Machine learning has shown to be effective at improving prediction accuracy on stock market securities or crude oil pricing \cite{jain1996, YU20082623, zhang2009stock, thawornwong2004forecasting}, especially using recurrent- (RNN) or artificial neural networks (ANN). % \cite{crotty2009} \cite{garber1989}
Having established the relationship between stock market values and economic development, the question remains whether the pandemic actually influenced stock market values. Social and economic factors have shown historical importance in stock market crashes, such as with the Tulip Mania, the Wall Street Crash of 1929 and the Great Depression \cite{white1990, kindleberger1986world} and the global financial crisis of 2008. An important topic during the recent pandemic was whether it would influence the economy and what its impact would be. Historically pandemics have shown their effects \cite{osterholm2017preparing, correia1918pandemics, jorda2020longer}, but the quantitative relationship between the COVID-19 pandemic and economic development is still an object of research \cite{chudik2020economic, baldwin2020economics, fernandes2020economic, deb2020economic}. During the pandemic daily data was registered on its development, such as infection rates, deaths, vaccinations and government restrictions \cite{dong2020interactive, hale2020variation}. As stock market crashes generally are a social phenomenon, this research investigates the relationship between these features and stock market pricing. The following research questions are posed:
\begin{itemize}
    \item[RQ1] \ldots To what extent does the COVID-19 pandemic data influence stock market values?
        \begin{itemize}
            \item What is the baseline performance for standard statistical models on short term and long term?
            \item How do different statistical and machine learning models compare given the same data?
            \item How do models generalize to different time-periods, such as distinctions between different infection waves?
            \item Which variables in a pandemic affect stock market pricing and to what extent? % Variables such as ...
            % \begin{itemize}
            %     \item COVID-19 pandemic data incl.: infections, deaths, vaccinations and % government restrictions
            %     \item Environmental factors such as calendar data or weather
            % \end{itemize}
        \end{itemize}
    \item[RQ2] \ldots To what extent can machine learning techniques aid in predicting stock market pricing, given a comparison of baseline models and the addition of pandemic data?
\end{itemize}
As the pandemic influenced a wide array of human behaviours, variables are restricted to the ones mentioned above. Other variables, such as mobility, energy consumption and social contacts have shown large differences during the pandemic, but are outside of the scope of this research.
% \paragraph{Overview of paper}
% To create context and support of the hypothesis, this paper first discusses related work \ref{sec:rel}. With this context, a proper methodology \ref{sec:meth} is formulated with an explanation on the data collection \ref{sec:data}, model development \ref{sec:models} and how these answer the proposed research questions \ref{sec:eval}. Using the methodology and proper context from related work, results can be analysed \ref{sec:res}, discussed \ref{sec:dis} and be made to construct a conclusion \ref{sec:conc}.